Red States = Red Flags?

### 2024-02-15 paper/paper.qmd

# Workspace Set up
library(tidyverse)
library(haven)
library(tidyr)
library(readxl)
library(dplyr)
library(readr)
library(knitr)
library(ggplot2)
library(lmtest)
library(sandwich)
```

@@

Where:

@@

@@

Discussion \# Medicaid - remove heading after The Affordable Care
(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to Figure C1, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In Figure C3, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In Figure C2,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

@@

Second discussion: Political affiliation

```{r}
# POLITICAL AFFILIATION MAP
```

>>>>

### Preparing Workspace
# Install and load necessary packages
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("knitr", quietly = TRUE)) install.packages("knitr")
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")
if (!requireNamespace("haven", quietly = TRUE)) install.packages("haven")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
if (!requireNamespace("lmtest", quietly = TRUE)) install.packages("lmtest")
if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")

# Workspace Set up
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(readr)
library(maps)
library(haven)
library(readxl)
library(here)
library(lmtest)
library(sandwich)
```

@@

Model set-up
=======
Where:

@@

```{r echo=FALSE, message=FALSE, results='asis'}
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
```

The range of birth rates in 2001 spans from approximately 17 per 1000
(excluding Utah (UT)) people in certain states to below 15 per 1000 in
others, with a convergence of rates around 12 to 14 per 1000.

The states in this graph are ones which had a policy change from no
delay to delay which allows for the analysis of the before-and-after
effects of policy implementation.

For example, Arizona's (AZ) implementation year of 2009 shows birth rate
descends sharply afterward. In contrast, North Carolina (NC) in 2011
didn't show such an immediate shift.

The states without delay policies showcase a natural decline in birth
rates over time, without the potential influence of policy changes. This
trend is consistent across the states, with some variations in the
steepness of the decline.

For example, California (CA) maintains a relatively high birth rate over
the years, starting near 15 per 1000 in 2001 and ending around 12 per
1000 in 2019. This could be influenced by its socio-economic profile and
immigration patterns.

All states show a general decline in relative birth rates over the
19-year period, which is in line with the national trend of declining
birth rates.

State-Specific Trends:

California (CA): The trend in California shows a relatively steady
decline throughout the period, without any sharp changes.

Florida (FL): Florida's trend is similar to California's, with a gradual
decline.

North Carolina (NC): North Carolina, with a policy change in 201, shows
a relatively stable trend before the policy change and a sharper decline
after.

Texas (TX): Texas exhibited a noticeable decline after the policy
implementation year of 2003.

When contrasting the birth rates of Texas and California, two states
with large populations, we see that Texas's birth rates declined more
sharply after its policy change in 2003 compared to California. This
might suggest an association between the policy implementation and the
accelerated decline in birth rates in Texas.

Comparing Georgia and North Carolina, both of which implemented delay
policies. Georgia's birth rates were rising before the policy change and
then started declining after, while North Carolina's birth rates were
relatively stable before 2011 and then declined.

```{r echo=FALSE, message=FALSE}
#Education

# Load Education data
educ_data <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")

# Creating the plot
ggplot(educ_data, aes(x = year)) +
  geom_line(aes(y = numbirth_lesshs, color = "Less than HS"), linewidth = 1) +
  geom_line(aes(y = numbirth_hsgrad, color = "HS Graduate"), linewidth = 1) +
  geom_line(aes(y = numbirth_somecoll, color = "Some College"), linewidth = 1) +
  geom_line(aes(y = numbirth_colgrad, color = "College Graduate"), linewidth = 1) +
  geom_line(
    aes(y = numbirthalleduc, color = "All Education Levels"),
    size = 1,
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = c(
      "Less than HS" = "blue",
      "HS Graduate" = "green",
      "Some College" = "orange",
      "College Graduate" = "red",
      "All Education Levels" = "grey"
      )
  ) +
  labs(
    title = "Number of Births by Education Level Over Time",
    x = "Year",
    y = "Number of Births",
    color = "Education Level"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Display the plot
ggsave("births_by_education_level_over_time.png",
       width = 10,
       height = 6)

# State Political Affiliation and Birth Rates
```{r echo=FALSE, message=FALSE}
# https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text

#reading the UKNCPR Datat
poldata <- read_csv(here("data", "analysis_data", "UKCPR_National_Welfare_Data_Update_020623.csv"))

poldata <- poldata %>%
  select(state_name,
         year,
         Population,
         `Fraction of State Senate that is Democrat`)

# Create a mapping of state abbreviations to full names
state_abbrevs <- state.abb
state_names <- state.name

# Create a data frame for the mapping
state_mapping <-
  data.frame(abbrev = tolower(state_abbrevs),
             name = tolower(state_names))

#Get numbirths data
numbirths_data <- read_csv(here("data", "analysis_data", "annual_policy", "numbirths_2001_2019.csv"))
births_2019 <- numbirths_data %>% filter(year == 2019)

# Convert state abbreviations to full names in your datasets
numbirths_2019_fullname <- births_2019 %>%
  mutate(stname = tolower(stname)) %>%
  left_join(state_mapping, by = c("stname" = "abbrev")) %>%
  select(-stname) %>%
  rename(stname = name)

poldata_2019 <- poldata %>% filter(year == 2019)

poldata_2019_fullname <- poldata_2019 %>%
  mutate(state_name = tolower(state_name)) %>%
  left_join(state_mapping, by = c("state_name" = "abbrev")) %>%
  select(-state_name) %>%
  rename(state_name = name)

# Merge poldata with numbirths_data using full state names
data_merged <-
  left_join(numbirths_2019_fullname,
            poldata_2019_fullname,
            by = c("stname" = "state_name"))

states_map <- map_data("state")

# Merge with states map data
map_data_merged <-
  left_join(states_map, data_merged, by = c("region" = "stname"))

# Plot
ggplot(map_data_merged) +
  geom_polygon(
    aes(
      x = long,
      y = lat,
      group = group,
      fill = `Fraction of State Senate that is Democrat`
    ),
    color = "white"
  ) +
  geom_text(
    data = map_data_merged %>% group_by(region) %>% summarise(
      long = mean(long),
      lat = mean(lat),
      numbirths = first(numbirth1544)
    ),
    aes(x = long, y = lat, label = numbirths),
    size = 3,
    check_overlap = TRUE
  ) +
  scale_fill_gradient(low = "red",
                      high = "blue",
                      name = "Democrat Fraction") +
  coord_fixed(1.3) +
  labs(title = "2019 Births and Democrat Fraction in State Senate by State") +
  theme_void() +
  theme(legend.position = "right")
```

Discussion

## First discussion point: Abortion

#relative average \# of births in states delay

```{r echo=FALSE, message=FALSE}
# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%

@@

### Discussion

# Medicaid - remove heading after The Affordable Care
(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to Figure C1, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In Figure C3, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In Figure C2,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

@@

Second discussion: Political affiliation and birth rates

From our analysis of political affiliations over the years, there are some relation between the two, especially in years following aggresive rhetorics or policies about reproductive rights. This indicates that periods following aggressive political rhetoric or significant policy changes regarding reproductive rights often coincide with noticeable changes in birth rates. This trend is observable across various states and is particularly pronounced in areas where political affiliation (# of democrats in state senate) change very differently from one side to the other. The rhetoric employed by political leaders and the nature of the policies they implement can have a substantial impact on individuals' choices regarding family planning. For instance, policies that limit access to reproductive healthcare, including contraception and abortion services, can lead to an increase in birth rates, especially in regions where alternative options are scarce.[@Nargund2009DecliningBirthRate]

Conversely, political movements that advocate for and expand access to reproductive healthcare are often associated with a stabilization or decrease in birth rates. This can be attributed to increased autonomy over reproductive choices, allowing individuals to make informed decisions about if and when to have children.
